---
title: "Analysis of Gene Expression"
subtitle: "Transcriptional signature of prion-induced neurotoxicity in a Drosophila model of transmissible mammalian prion disease."
author: Niek Scholten
date: "`r Sys.Date()`"
output: pdf_document
---

[//]: # (Page 1)
\thispagestyle{empty}

\vspace{100pt}
\begin{center}
    \includegraphics[width=5in,height=5in]{drosophila.jpg}\LARGE\\
\end{center}
\setcounter{secnumdepth}{2}

[//]: # (Page 2)
\newpage


\clearpage
\setcounter{page}{1}

\tableofcontents

[//]: # (Page 3)
\newpage


# Setup
```{r setup, echo = T, results = 'hide', message = F, warning = F}
# Options for all chunks
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)

# Load the packages & register the amount of workers
packages <- c("affy", "scales", "DESeq2", "BiocParallel",
              "pheatmap", "PoiClaClu", "ggplot2", "knitr", "pander")
invisible(lapply(packages, library, character.only = TRUE))
register(MulticoreParam(12))

# Load the data into a data frame
data <- read.table("Data/GSE144028.txt")

# Define groups for the replicants
group <- c("X51D_5_NBH",
            "X51D_5_S",
            "X51D_30_NBH",
            "X51D_30_S",
            "PrPCyt_5_NBH",
            "PrPCyt_5_S",
            "PrPCyt_30_NBH",
            "PrPCyt_30_S",
            "PrPGPI_5_NBH",
            "PrPGPI_5_S",
            "PrPGPI_40_NBH",
            "PrPGPI_40_S")
groups <- factor(rep(1:12, each=3),
                 labels = group)

# Set color distributions for the graphs
colors12 <- hue_pal()(12)
colors36 <- rep(colors12, each=3)
```
This is the setup of the project. It loads all the necessary packages and sets values that are important for later.

[//]: # (Page 4)
\newpage

# Initial analysis
The initial analysis includes a summary of the data and a quick look at the visualisation of this data in a boxplot.

## Summary
```{r}
sumMatrix <- summary(data)
sumFrame <- data.frame()
sumGroups <- split(sumMatrix, ceiling(seq_along(sumMatrix)/6))
sumGroups <- lapply(sumGroups, function (i) {
  for (j in i){
      as.numeric(
        strsplit(j, ":")[[1]][1][2])
                }
            }
       )
# for (sumGroup in sumGroups) {
#     sumFrame <- rbind(sumFrame, as.numeric(strsplit(sumGroup, ":")[[1]][2]))
# }


# Disable intertable text
panderOptions('table.continues', '')
# Pretty print the output of the data summary
pander(summary(data), split.tables = 64)
```
[//]: # (Page 5)
\newpage


## Boxplot
```{r boxplot, fig.width = 10, fig.height = 8}
# Create a boxplot for initial analysis
boxplot(log2(data+0.1),
        outline = FALSE,
        col = colors36,
        horizontal = TRUE,
        las = 2,
        main = "Distrubution of count values",
        cex.axis= 0.6)
```

```{r densityplot}
myColors <- hue_pal()(12)

plotDensity(log2(data + 0.1), col=colors36,
            lty= seq_len(ncol(data)), xlab="Log2(count)",
            main="Expression Distribution")

legend('topright', names(data), lty= seq_len(ncol(data)),
       col=colors36,
       cex=0.32)  # Fix scale for knitted output
abline(v=-1.5, lwd=1, col='red', lty=2)
```
```{r DESeq}
(ddsMat <- DESeqDataSetFromMatrix(countData = data,
                                  colData = data.frame(samples = names(data)),
                                  design = ~ 1))
rld.dds <- vst(ddsMat)
rld <- assay(rld.dds)

sampledists <- dist( t( rld ))
```
```{r heatmap}
sampleDistMatrix <- as.matrix(sampledists)

annotation <- data.frame(Type = factor(rep(rep(1:2, each = 3), each = 6),
                                          labels = c("Normal Brain Homogenate",
                                                     "Scrapie")))

rownames(annotation) <- names(counts)

pheatmap(sampleDistMatrix, show_colnames = FALSE,
         # annotation_col = annotation,  # Gives an error
         clustering_distance_rows = sampledists,
         clustering_distance_cols = sampledists,
         main = "Euclidian Sample Distances")
```
```{r MDS}
dds <- assay(ddsMat)
poisd <- PoissonDistance( t(dds) )

samplePoisDistMatrix <- as.matrix(poisd$dd)

mdsPoisData <- data.frame( cmdscale(samplePoisDistMatrix) )

names(mdsPoisData) <- c('x_coord', 'y_coord')

coldata <- names(data)

ggplot(mdsPoisData, aes(x_coord, y_coord, color = groups, label = coldata)) +
  geom_text(size = 4) +
  ggtitle('Multi Dimensional Scaling') +
  labs(x = "Poisson Distance", y = "Poisson Distance") +
  theme_bw()
```
Some samples clearly deviate from the other 2 in the group.
This is especially clear with X51D_5_S, PrPCyt_30_NBH, X51D_30_NBH & PrPCyt_5_S.
Strangely, these samples are all the first one in their respective group. This could indicate that the first tests were less accurate.
Since 3 samples must remain in each group, no data will be removed from the set.
```{r}
counts.fpm <- log2( fpm(ddsMat, robust = TRUE) + 1 )
dds <- DESeq(ddsMat, parallel = TRUE)
res <- results(dds)
```
```{r}
beforeCounts <- counts(dds)
keep <- rowSums(beforeCounts) >= 10
dds <- dds[keep,]
afterCounts <- counts(dds)

countCompare <- data.frame(nrow(beforeCounts),
                           nrow(afterCounts),
                           nrow(beforeCounts) - nrow(afterCounts))
colnames(countCompare) <- c("Counts before filtering",
                            "Counts after filtering",
                            "Difference in counts")
kable(countCompare)
```
The dataset has been trimmed to filter out genes with count values lower than 10.
This results in a smaller dataset because more than 4000 genes have been removed.
